Practical Optional Types for Clojure

### 2015-12-03 esop-rebuttal.txt

Major issues:

1) Section 5.1 should use tables instead of prose.

We will restate the data in tables.

==========================================================================

Q: What is the delta between occurrence typing in ICFP'10 and the as presented here?

A:

Q: Is TC decidable?

Q: Why can't the type system check Example 9? What restrictions are imposed on HMaps?

A: For presentational reasons, we only support a specific pattern of looking up
a known key in a union of HMaps...

Q: What specific patterns could not be supported by TC that led to CircleCI
   abandoning it?

Q: How is TC compiled/used in practice? Is it sealed from untyped namespaces?

A: core.typed

Q: "Java cannot be passed nil"?

A: This is a conservative assumption, as the paper states in the same sentence.
   In practice this is a reasonable default, as discussed in the CircleCI evaluation (5.1).

Q: Does (dissoc k) just remove its argument-entry from a map?

A: (dissoc m k) returns a new map that is like m, but without a k entry.

Q: How symbolic representations of objects in OT relate to Scala's
   path-dependent types?

A: ??

Q: Formula (3) -- do the positive/negative statements mean that d
   is of type Nat, so N_d should be read as a proposition?

A: Yes, N_d says d is a Nat, \overline{N_d} says d is not a Nat.

Q: Are Java's generic classes (in particular collections) supported
   in TC type checking algorithm?

A: No. We chose to avoid the added complexity of Java generics ..

Q: How do Clojure IDE's "incrementalize" type checking for code completion?

A: Cursive Clojure prioritizes type hints assigned to variables in the local scope
   during method name autocompletion.

Q: "Then (new[S] F "a/b") is checked against a nilable conversion..."?

A: FIXME I have no idea, looks like a bug?

Q: What is the type of (defmethod e_m e_v e_f)?

A: As shown in T-DefMethod, it returns the same type as e_m.

Q: How overlapping multi-method clauses are handled?

A: In the type system, each method is checked in isolation with respect to 
   just the current dispatch value and the dispatch function type.
   In the operational semantics, B-BetaMulti can only fire if there is an unambiguous
   match, encoded in the GM metafunction.

Q: Citation for Footnote 1?

A: FIXME

Q: What category do you consider e.g. C# 4.0, or Scala's Dynamic, or
   Dart? (WRT gradual typing)

Q: Is the additional of an Unknown or Dynamic type in a static
    type system enough to make it gradual?

Q: Is "optional" here a synonym for Bracha's "pluggable" type system?

A: Yes, because the type system does not affect the final program.
  ... But also no...

Q: One way of thinking about those
   annotations is that in fact Typed Clojure actually is a gradual type
   system as well as an optional type system, with the annotations
   effectively acting as type assumptions which will be checked
   dynamically to interact with other untyped code.

A: Typed Clojure does not check these assumptions at runtime, thus is not
   gradually typed in any sense, just optionally typed like TypeScript.

Q: What is "ann"?

A: (ann v t) 'ann'otates variable v as type t. See examples in Section 2.2.

Q: What is the "Any" type?

A: "Any" is effectively the same as (U nil Object), and has no special runtime semantics.
   This also holds in the implementation, as JVM primitives pun their boxed equivalents
   as static types, reflecting how Clojure deals with primitives.

Q: Why does 42 have the object "0"?

A: There is no known way to derive 42 from the current environment, so the object is "empty".
   More recent forumlations of occurrence typing assign the object "42" to aid symbolic
   computations.

Q: Is the complexity of multimethods worth it, given the case studies?

A:

### 2015-12-04 esop-rebuttal.txt

We thank the reviewers for their detailed feedback.  We are thrilled the
reviewers found the paper "a joy to read", "an excellent demonstration 
of occurrence typing's power", and providing a "substantial practical 
contribution".

The reviewers' primary concerns are with two issues:

1. What are the contributions beyond occurrence typing as presented
   in ICFP'10?

There are two main extensions in relation to occurrence typing.

The first is the presentation of HMap's, with support for flow checking
on sub-parts of maps. This required several extra typing rules (Fig. 10)
and subtle, but clean, extensions to the `update` rule (Fig. 12, clauses
4-7).

The second is multimethods, which is mainly contained in the T-DefMethod
typing rule. It is in the same spirit as T-If, which was present in ICFP'10,
but demonstrates occurrence typing extends gracefully to support complex 
control flow patterns.

The main takeaway is that occurrence typing as formulated in ICFP'10
serves as a valuable base to support for the nuances of a Clojure type system,
without any particular foresight from the original authors.

2. Are multimethods worth the extra complexity in the type system, given
   they are used so little in practice?

Yes.

The numbers presented in the evaluation section are possibly misleading.
Even though a low number of multimethods are defined in practice, extending
them via `defmethod` accounts for a large amount of code.

A comparable analogy to a Java code base might be assuming interfaces are
a minor feature if a small number of interfaces are defined, yet a large number
of classes extend them.

As noted in Section 5.1, these multimethods are single-dispatch on a
fixed map entry, which further gives them similar roles as Java interfaces,
which perform single dispatch on the first argument of methods.

Following are responses to specific statements or queries.

@@

==========================================================================

Q. Section 5.1 should use tables instead of prose.

A. We will restate the data in tables.

Q: Is TC decidable? `restrict` seems problematic.

A: The calculus leaves the details of `restrict` up to implementations. core.typed
   features an ad-hoc intersection type, so `restrict` simply returns the intersection 
   of both types, thus making the occurrence typing implementation decidable.

@@

Q: Why can't the type system check Example 9? What restrictions are imposed on HMaps?

A: For presentational reasons we only support a specific pattern of looking up
   a common, known key in a union of HMaps (notice T-GetHMap (Fig. 10) is the only rule
   supporting lookups on unions of HMaps). Example 9 is checkable in core.typed, 
   but not the calculus, and serves as a demonstration of the kinds of patterns
   that might occur in Clojure code.

@@

Q: What specific patterns could not be supported by TC that led to CircleCI
   abandoning it?

A: We refer to the detailed blog post they published:

     http://blog.circleci.com/why-were-no-longer-using-core-typed/

Q: Are the case studies representative programs that would be good candidates for optiona
   typing?

A: We chose the CircleCI code base as it was a substantial size. The other case
   study was a randomly chosen open source project. A goal of Typed Clojure is to support
   as many Clojure idioms as possible, and we learnt valuable lessons from both projects
   in this regard.

@@

Q: How is TC compiled/used in practice? Is it sealed from untyped namespaces?

A: TC is simply compiled to be normal Clojure code. As alluded to in Footnote 1,
   Typed Clojure is not gradually typed.

@@

A: This is a conservative assumption, as the paper states in the same sentence.
   In practice this is a reasonable default, as discussed in the CircleCI evaluation (5.1).

Q: Does (dissoc k) just remove its argument-entry from a map?

A: (dissoc m k) returns a new map that is like m, but without a k entry.

Q: How do symbolic representations of objects in OT relate to Scala's
   path-dependent types?

A: They are unrelated.

@@

Q: Formula (3) -- do the positive/negative statements mean that d
   is of type Nat, so N_d should be read as a proposition?

A: Yes, N_d says d is a Nat, \overline{N_d} says d is not a Nat.

Q: Are Java's generic classes (in particular collections) supported
   in TC type checking algorithm?

A: No. TC however supports parametric polymorphism for Clojure collections and a manual subset
   of Java collections. In practice, this is sufficient for many interop situations.

Q: The simplified syntax chosen for method calls works for compilers, but
   not IDE's. How do Clojure IDE's "incrementalize" type checking for code completion?

A: This syntax is not arbitrary; (.method o) is sugar for (. o (method)) in Clojure.
   Cursive Clojure prioritizes type hints assigned to variables in the local scope
   during method name autocompletion.

@@

Q: How overlapping multi-method clauses are handled?

A: In the type system, each method is checked in isolation with respect to 
   just the current dispatch value and the dispatch function type.
   In the operational semantics, B-BetaMulti can only fire if there is an unambiguous
   match, encoded in the GM metafunction.

Q: Citation for Footnote 1?

A: Siek et al SNAPL'15 is a good candidate for describing gradual typing, however optional
   typing is not explicitly defined.

@@

Q: What category do you consider e.g. C# 4.0, or Scala's Dynamic, or
   Dart? (WRT gradual typing)

A: C# 4.0 is gradually typed. Scala's Dynamic is an unrelated metaprogramming construct, 
   comparable to a statically resolved "method_missing" in Ruby.  Dart is 
   optionally typed.

@@

Q: Is the additional of an Unknown or Dynamic type in a static
   type system enough to make it gradual?

A: A gradual type system is defined by its operational semantics, in particular
   how the runtime prevents untyped code from breaking static invariants.

@@

Q: One way of thinking about those annotations is that in fact Typed Clojure 
   actually is a gradual type system as well as an optional type system, with 
   the annotations effectively acting as type assumptions which will be checked
   dynamically to interact with other untyped code.

A: Typed Clojure does not check these assumptions at runtime, thus is not
   gradually typed in any sense, just optionally typed like TypeScript.

Q: What is "ann"?

A: (ann v t) 'ann'otates variable v as type t. See examples in Section 2.2.
   We will explain this explicitly.

Q: What is the "Any" type? "Kw"?

A: "Any" is effectively the same as (U nil Object), and has no special runtime semantics.
   It is *not* a Dynamic type.
   "Kw" is the type for keywords.

@@

Q: Why does 42 have the object "0"?

A: There is no known way to derive 42 from the current environment, so we say the 
   object is "empty".

### 2015-12-04 esop-rebuttal.txt

We thank the reviewers for their detailed feedback.  We are thrilled the
reviewers found the paper "a joy to read", "an excellent demonstration 
of occurrence typing's power", and providing a "substantial practical 
contribution".

The reviewers' primary concerns are with three issues:

1. What are the contributions beyond occurrence typing as presented
   in ICFP'10?

As described in Section 1, our overall contributions are

  - a type system for Clojure that support Clojure idioms,
  - a sound formal model for three crucial type system features, and
  - an evaluation of the type system in practice.

There are two main extensions specifically in relation to occurrence typing.

@@

The numbers presented in the evaluation section are possibly misleading.
Even though a low number of multimethods are defined in practice, extending
them via `defmethod` accounts for a large amount of code.

A comparable analogy to a Java code base might be assuming interfaces are
a minor feature if a small number of interfaces are defined, yet a large number
of classes extend them.

As noted in Section 5.1, these multimethods are single-dispatch on a
fixed map entry, which further gives them similar roles as Java interfaces,
which perform single dispatch on the first argument of methods.

3. What is the distinction between optional and gradual typing?

Optional typing refers to type systems that do not enforce type invariants
at runtime, and are comparable to Bracha's "pluggable" type systems (as
suggested by Review 2).

Typed Clojure does not check assumptions at runtime, thus is not
gradually typed in any sense, just optionally typed like TypeScript.

Review 2 requests to categorize the following:
 - C# 4.0 is gradually typed. 
 - Scala's Dynamic is an unrelated metaprogramming construct, 
   comparable to a statically resolved "method_missing" in Ruby. 
 - Dart is optionally typed.

A gradual type system is defined by its operational semantics, in particular
how the runtime prevents untyped code from breaking static invariants. An
addition of an "Unknown or Dynamic type" (Review 2) is not enough to make
a gradual type system.

We distinguish optional and gradual typing in footnote 1. A good reference for 
this distinction is Siek et al SNAPL'15 (as requested by Review 2).

"Any" is effectively the same as `(U nil Object)`, and has no special runtime semantics.
It is *not* a Dynamic type (Review 2).

@@

==========================================================================

Following are responses to specific statements or queries.

REVIEW 1: Section 5.1 should use tables instead of prose.

  We will restate the data in tables.

REVIEW 1: Is TC decidable? `restrict` seems problematic.

  The calculus leaves the details of `restrict` up to implementations. core.typed
  features an ad-hoc intersection type, so `restrict` simply returns the intersection 
  of both types, thus making the occurrence typing implementation decidable.

REVIEW 1: Why can't the type system check Example 9? 
          What restrictions are imposed on HMaps?

  For presentational reasons we only support a specific pattern of looking up
  a common, known key in a union of HMaps (notice T-GetHMap (Fig. 10) is the only rule
  supporting lookups on unions of HMaps). Example 9 is checkable in core.typed, 
  but not the calculus, and serves as a demonstration of the kinds of patterns
  that might occur in Clojure code.

REVIEW 1: What specific patterns could not be supported by TC that led to CircleCI
          abandoning it?

  We refer to the detailed blog post they published:

@@

    http://blog.circleci.com/why-were-no-longer-using-core-typed/

REVIEW 3: Are the case studies representative programs that would be good candidates 
          for optional typing?

  We chose the CircleCI code base as it was a substantial size. The other case
  study was a randomly chosen open source project. A goal of Typed Clojure is to support
  as many Clojure idioms as possible, and we learnt valuable lessons from both projects
  in this regard.

REVIEW 1: How is TC compiled/used in practice? Is it sealed from untyped namespaces?

  TC is compiled to as normal Clojure code. As alluded to in Footnote 1,
  Typed Clojure is not gradually typed.

REVIEW 1: "Java cannot be passed nil"?

  This is a conservative assumption, as the paper states in the same sentence.
  In practice this is a reasonable default, as discussed in the CircleCI evaluation (5.1).

REVIEW 1: Does (dissoc k) just remove its argument-entry from a map?

  (dissoc m k) returns a new map that is like m, but without a k entry.

REVIEW 1: How do symbolic representations of objects in OT relate to Scala's
          path-dependent types?

  They are unrelated.

REVIEW 1: Formula (3) -- do the positive/negative statements mean that d
          is of type Nat, so N_d should be read as a proposition?

  Yes, N_d says d is a Nat, \overline{N_d} says d is not a Nat.

REVIEW 1: Are Java's generic classes (in particular collections) supported
          in TC type checking algorithm?

  No. TC however supports parametric polymorphism for Clojure collections and a manual subset
  of Java collections. In practice, this is sufficient for many interop situations.

REVIEW 1: The simplified syntax chosen for method calls works for compilers, but
          not IDE's. How do Clojure IDE's "incrementalize" type checking for code completion?

  This syntax is not arbitrary; (.method o) is sugar for (. o (method)) in Clojure.
  Cursive Clojure prioritizes type hints assigned to variables in the local scope
  during method name autocompletion.

REVIEW 1: "Then (new[S] F "a/b") is checked against a nilable conversion..."?

  The prose should be fixed to instead read "non-nilable".

REVIEW 1: What is the type of (defmethod e_m e_v e_f)?

  As shown in T-DefMethod, it returns the same type as e_m.

REVIEW 1: How overlapping multi-method clauses are handled?

  In the type system, each method is checked in isolation with respect to 
  just the current dispatch value and the dispatch function type.
  In the operational semantics, B-BetaMulti can only fire if there is an unambiguous
  match, encoded in the GM metafunction.

REVIEW 2: What is "ann"?

  (ann v t) 'ann'otates variable v as type t. See examples in Section 2.2.
  We will explain this explicitly.

REVIEW 2: What is the "Kw" type?

   "Kw" is the type for keywords.

REVIEW 2: Why does 42 have the object "0"?

  There is no known way to derive 42 from the current environment, so we say the 
  object is "empty".

### 2015-12-04 esop-rebuttal.txt

The reviewers' primary concerns are with three issues:

1. What are the contributions beyond occurrence typing as presented
   in ICFP'10?

As described in Section 1, our overall contributions are

  - a type system for Clojure that support Clojure idioms,
  - a sound formal model for three crucial type system features,
  - an implementation successfully used in practice and industry, and
  - an evaluation of the type system in practice.

### 2015-12-04 esop-rebuttal.txt

We distinguish optional and gradual typing in footnote 1. A good reference for 
this distinction is Siek et al SNAPL'15 (as requested by Review 2).

"Any" is effectively the same as `(U nil Object)`, and has no special runtime semantics.
It is *not* a Dynamic type (Review 2).

==========================================================================

We now address a few specific questions from the reviewers. The
reviewers also make several suggestions for presentation revisions and
improvements to the evaluation, including comparisons to other systems---we 
will make these changes in a revised version.

### 2015-12-05 esop-rebuttal.txt

We thank the reviewers for their detailed feedback.  We're happy the
reviewers found the paper "a joy to read", "an excellent demonstration 
of occurrence typing's power", and provided "substantial practical 
contribution".

We begin with three high-level issues:

@@

1. What are the contributions beyond occurrence typing as presented
   in ICFP'10?

As described in Section 1, our overall contributions are

  - a type system for Clojure supporting Clojure idioms,
  - a sound formal model for three crucial type system features,
  - an implementation successfully used in practice and industry, and
  - an evaluation of the type system in practice.

There are two main extensions specifically related to occurrence typing.

The first is the presentation of HMap's, with support for flow
checking on sub-parts of maps. This required several extra typing
rules (Fig. 10) and subtle extensions to the `update` rule (Fig. 12,
clauses 4-7).

The second is multimethods, which is mainly contained in the
T-DefMethod typing rule. It reuses the machinery T-If, which was
present in ICFP'10, but demonstrates occurrence typing is able to
support new and previously-unhandled control flow.

2. Are multimethods worth the extra complexity in the type system, if
   they are used rarely?

@@

Yes.

The numbers presented in the evaluation section do not indicate that
multimethods are are.  Even though a low number of distinct
multimethods are defined in practice, extending them via `defmethod`
accounts for a large amount of code.

Analagously, though Java code may define few interfaces, they will be
used in more class definitions, and then in many method calls.

As noted in Section 5.1, these multimethods are often single-dispatch
on a fixed map entry, making them even more similar to Java
interfaces, which perform single dispatch on the first argument of
methods.

@@

3. What is the distinction between optional and gradual typing?

Optional typing refers to type systems that do not enforce type invariants
at runtime, and are comparable to Bracha's "pluggable" type systems (as
suggested by Review 2).

Typed Clojure does not check assumptions at runtime, thus is not
gradually typed, just optionally typed like TypeScript.

Review 2 requests to categorize the following:
 - C# 4.0: gradually typed. 
 - Scala's Dynamic: a mostly-unrelated metaprogramming construct, 
   comparable to a statically resolved "method_missing" in Ruby. 
 - Dart: optionally typed.

A gradual type system is defined by its operational semantics, in particular
that the runtime prevents untyped code from breaking static invariants. An
addition of an "Unknown or Dynamic type" (Review 2) is not enough to make
a gradual type system without this enforcement.

We distinguish optional and gradual typing in footnote 1; see [Siek et
al, SNAPL'15] for more on the distinction.

"Any" is the top type, effectively the same as `(U nil Object)`, and
has no special runtime semantics.  It is *not* a Dynamic type (Review
2).

@@

==========================================================================

We now address a few specific questions from the reviewers. The
reviewers also make several suggestions for presentation revisions and
improvements to the evaluation, including comparisons to other systems---we 
will make these changes in a revised version.

REVIEW 1: Is TC decidable? `restrict` seems problematic.

  The calculus leaves the details of `restrict` up to
  implementations. core.typed features an ad-hoc intersection type, so
  `restrict` simply returns the intersection of both types, thus
  making this function simple.

@@

REVIEW 1: Why can't the type system check Example 9? 
          What restrictions are imposed on HMaps?

  For presentational reasons we only support a specific pattern of
  looking up a common, known key in a union of HMaps (notice T-GetHMap
  (Fig. 10) is the only rule supporting lookups on unions of
  HMaps). Example 9 is checkable in our implementation, but not the
  calculus, and serves as a demonstration of the kinds of patterns
  that might occur in Clojure code.

@@

REVIEW 1: Formula (3) -- do the positive/negative statements mean that d
          is of type Nat, so N_d should be read as a proposition?

  Yes, N_d says d is a Nat, \overline{N_d} says d is not a Nat.

REVIEW 1: Are Java's generic classes (in particular collections) supported
          in TC type checking algorithm?

  No. TC however supports parametric polymorphism for Clojure
  collections and a manual subset of Java collections. In practice,
  this is sufficient for most interop situations.

@@

REVIEW 1: The simplified syntax chosen for method calls works for
          compilers, but not IDE's. How do Clojure IDE's
          "incrementalize" type checking for code completion?

  This is not our choice of syntax; (.method o) is sugar for (. o
  (method)) in Clojure.  The Cursive Clojure IDE prioritizes type
  hints assigned to variables in the local scope during method name
  autocompletion.

@@

  The prose should be fixed to instead read "non-nilable".

REVIEW 1: What is the type of (defmethod e_m e_v e_f)?

  As shown in T-DefMethod, it returns the same type as e_m.

REVIEW 1: How are overlapping multi-method clauses handled?

  In the type system, each method is checked in isolation with respect
  to the current dispatch value and the dispatch function type.  In
  the operational semantics, B-BetaMulti can only fire if there is an
  unambiguous match, encoded in the GM metafunction.

@@

REVIEW 2: What is the "Kw" type?

   "Kw" is the type for keywords.

REVIEW 2: Why does 42 have the object "0"?

  Objects only describe references to the runtime environment, so we
  say the object of 42 is "empty".

### 2015-12-26 formal-model.tex
fix italic gamma

We present Java interoperability in a restricted setting without class inheritance,
overloading or Java Generics.

We extend the syntax in \figref{main:figure:javatyping} with Java field lookups and calls to
methods and constructors. 
To prevent ambiguity, % between zero-argument methods and fields, 
field accesses are written \fieldexp{\fld{}}{\e{}}
and method calls $\methodexp{\mth{}}{\e{}}{\overrightarrow{e}}$.
%and \clj{(new class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.

@@

\paragraph{Methods} Next we convert {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}
to the non-reflective expression
{\methodstaticexp {\File} {} {\String} {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}.
%The T-Method rule checks unresolved methods.
%$$
%    {\TMethod}
%$$
We use T-Method to check {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}, 
which checks unresolved methods.
We verify the target type \s{} = \File is non-nil before erasing \Nil
by converting to a Java type \classhint{1} = \File.
The specific overload is chosen from the class table based on \classhint{1}---there is at most one.
Then {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}}
is checked against a nilable conversion of \classhint{1}, \t{1} = \Union{\Nil}{\File},
which succeeds by the previous rule.
The nilable return type \t{} = \Union{\Nil}{\String} is given, and---finally---the 
entire expression rewrites to expression \ref{eq:resolved}.
%
%We allow arguments to constructors and methods to be nilable, but not method
%and field targets.

The T-Field rule is included in \figref{main:figure:javatyping}, and is like T-Method, but without arguments.

@@

\figref{main:figure:mmsyntax}
presents \emph{immutable}
multimethods without default methods to ease presentation.
%Syntax and semantics are given in \figref{main:figure:mmsyntax}. 
%Multimethods can error if no matching method is chosen (rules in the supplemental material).
\figref{main:figure:mmexample}
is the translation of \egref{example:hi-multimethod} to \lambdatc{}.
%\begin{minted}{clojure}
%(ann hi [Kw -> Str])
%(defmulti hi identity)
%(defmethod hi :en [_] "hello")
%(defmethod hi :fr [_] "bonjour")
%(hi :en) ;=> "hello"
%\end{minted}

>>>>

We present Java interoperability in a restricted setting without class inheritance,
overloading or Java Generics.
%
We extend the syntax in \figref{main:figure:javatyping} with Java field lookups and calls to
methods and constructors. 
To prevent ambiguity between zero-argument methods and fields,
we use Clojure's primitive ``dot'' syntax:
field accesses are written \fieldexp{\fld{}}{\e{}}
and method calls $\methodexp{\mth{}}{\e{}}{\overrightarrow{e}}$.
%and \clj{(new class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.

@@

\paragraph{Methods} Next we convert {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}
to the non-reflective expression
{\methodstaticexp {\File} {} {\String} {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}.
%The T-Method rule checks unresolved methods.
%$$
%    {\TMethod}
%$$
The T-Method rule for unresolved methods
checks {\methodexp {\getparent{}} {\newstaticexp {\String} {\File{}} {\File{}} {\makestr{a/b}}} {}}.
We verify the target type \s{} = \File is non-nil by T-New.
The overload is chosen from the class table based on \classhint{1} = \File---there is at most one.
The nilable return type \t{} = \Union{\Nil}{\String} is given, and the 
entire expression rewrites to expression \ref{eq:resolved}.
%
%We allow arguments to constructors and methods to be nilable, but not method
%and field targets.

The T-Field rule (\figref{main:figure:javatyping}) is like T-Method, but without arguments.

@@

\figref{main:figure:mmsyntax} presents \emph{immutable} multimethods without default methods to ease presentation.
%Syntax and semantics are given in \figref{main:figure:mmsyntax}. 
%Multimethods can error if no matching method is chosen (rules in the supplemental material).
\figref{main:figure:mmexample} translates the mutable \egref{example:hi-multimethod} to \lambdatc{}.
%\begin{minted}{clojure}
%(ann hi [Kw -> Str])
%(defmulti hi identity)
%(defmethod hi :en [_] "hello")
%(defmethod hi :fr [_] "bonjour")
%(hi :en) ;=> "hello"
%\end{minted}
